A state space model reveals the rich and varied dynamics of hippocampal replay

### 2020-07-18 Article.tex
Draft 1

Mental events may not occur in the same time span as actual events: memories can be retrieved and used in less time than was required for the original experience, and mental simulations can span a series of actions more quickly than would be required to perform those actions. Consistent with this, generative representations can unfold rapidly. As a result, decoding generative representations often involves making assumptions about their timescale and temporal structure, and these assumptions can limit the interpretation of the data.

@@

Identifying these events typically involves multiple steps and assumptions about the nature of the event. First, an encoding model is constructed based on spiking activity during movement, most often using only spikes that have been clustered as single units (putative single neurons). Then, a subset of SWRs or events with high activity levels are selected based on an threshold for event size chosen by the experimenter (cite Foster and Wilson, Diba and Buzsaki, Karlsson and Frank, appropriate Csicsvari paper). The spikes within these events decode the probability of position in overlapping or non-overlapping time bins whose size is also chosen by the experimenter. Finally, the most commonly used approach to detecting sequential replay involves fitting a line to the resulting set of probability distributions, which instantiates the assumption that the representations progress at a constant speed (Foster and Wilson, Diba and Buzsaki, Karlsson and Frank,). A statistical test is then used to determining whether the fit is better than the fit to shuffled versions of the data.

While this approach identifies constant speed events, it does not consider events that are rejected by the statistical test, and the use of a fixed size temporal bin acts as a boxcar smoother that limits the potential movement speeds of the representation. The linear fit is also problematic, because it has the potential to reject real events that do not move at constant speeds. Approaches that focus on the order of cell activity within each even (Lee and Wilson, Gupta et. al?) allow for a relaxation of that linear assumption, but replace by either ignoring all but the first spike from each cell or some arbitrarily smoothed version of the spike train. These approaches also do not provide information about the dynamics of the underlying spatial representation. Moreover, there is now evidence that a subset of these events have stationary representations of a single location, which both approaches exclude. In addition, these methods identify only a subset of events as containing content (typically somewhere between 10 and 45\% of the chosen events), and thereby provide little insight into the remaining majority of the events.

@@

\section*{Results}
\begin{figure*}%[tbhp]
\centering
\includegraphics[width=0.80\linewidth]{figures/Figure1/Figure1_v4}
\caption{Components of the state space model and demonstration on simulated data. \textbf{(A)} 19 simulated place cells that exhibit three different sequence patterns. For the first third of time, one cell fires repeatedly, representing one stationary location. For the second third of time, the cells fire in sequence, representing a smooth continuous trajectory through space. For the last third cells fire in an incoherent pattern, representing a non-smooth fragmented trajectory through space. \textbf{(B)} Like the Bayesian Decoder, the state space model uses estimates of cells' place fields from when the animal is moving and combines them with the observed spikes in (A) to compute the likelihood of position for each time step. \textbf{(C)} This is combined with an explicit model of each dynamic state--which determines how latent position can change based on the posterior from the previous time step. In the stationary state, this is modeled as the identity matrix because the position remains the same. In the continuous state, the dynamics are modeled as a Gaussian random walk, meaning the next possible position is likely to be spatially close. In the fragmented state, the dynamics are modeled as uniform, meaning the next position is equally likely to be anywhere. Note that the dynamics displayed only reflect the within-state dynamics. \textbf{(D)} The probability of remaining in a particular state versus switching to another state. This is modeled as having a high probability of remaining in a particular state with a small probability of switching to one of the other dynamics at each time step. \textbf{(F)} The state space model uses the components in A-D over all time to decode the joint probability of latent position and dynamic state. The joint probability can be summarized by marginalizing over latent position (left panel) to get the probability of each dynamic state over time (lines). This can also be used to get an estimate of latent position over time (right panel). \textbf{(G)} The probability of each state given a particular constant speed trajectory. Each speed corresponds to a simulated spiking sequence at that speed. The lines correspond to the average probability of that state over the sequence. Dotted line represents the 0.8 classification threshold we use to classify each range of speeds and shaded regions correspond to the classification we have given the range of speeds.
}
\label{1}
\end{figure*}
\subsection*{Overview of the model}
We begin with an application of the model to simulated data, both to validate the model and to provide intuition. (Figure \ref{1}). We simulated 19 Poisson spiking cells with Gaussian place fields on a 180 cm virtual linear track. Each place field has a 36 cm variance and a 15 Hz peak firing rate, which is spaced every 10 cm along the virtual track. We then apply our decoding algorithm to the spiking sequence in Figure \ref{1}A. For the first third of this sequence, a single place cell fires repeatedly, representing a single location. For the middle third of the sequence, the cells fire in sequential spatial order, representing a fast moving trajectory. For the last third, the cells fire in a incoherent spatial order. These firing patterns represent three different types of sequence dynamics that could be expressed during replay events, which we call stationary, continuous, and fragmented, respectively. The goal of our model is to characterize SWRs in terms of a mixture of these three dynamics at every time point.

Decoding the spiking sequence dynamics requires specifying two elements: the data model---how the spikes relate to position---and the movement dynamics model--how the position can change over time. For the data model, our decoder is the same as the Bayesian decoder. We compute an estimate of how each cell's firing rate varies over position (the place field, Figure \ref{1}B). This is used during decoding to compute the Poisson likelihood of position over time given the spiking sequence of interest. In contrast to the Bayesian decoder, we can use arbitrarily small time bins (in our case 2 ms vs. 20 ms or more) because we are able to take advantage of the prior placed on the dynamics by the state space model. This allows us to detect changes on smaller time scales than would be possible with the Bayesian decoder.

Next, we specify a movement dynamics model for how latent position---the "mental" position of the animal represented by the cells--- evolves over time (Figure \ref{1}C). We do this by defining a movement state transition matrix, which defines how the latent position can change from the previous time step. Previous findings suggest three distinct dynamics: stationary \cite{JaiDistincthippocampalcorticalmemory2017, FarooqEmergencepreconfiguredplastic2019}, smoothly moving \cite{DavidsonHippocampalReplayExtended2009}, and fragmented / jumpy \cite{PfeifferAutoassociativedynamicsgeneration2015}, which could correspond to both spatially incoherent representations and representations that jump from one position to another. We therefore developed state transition models to capture each of these dynamics.

@@

Finally, we specify how likely each dynamic is to persist in time versus changing to another dynamic (Figure \ref{1}D). In order to be conservative, we assume that each dynamic is likely to dominate for ~100 ms, which is approximately the duration of the SWR. There is however,  a small probability of switching to one of the other dynamics (note that below we show that our results are relatively insensitive to this parameter). Accordingly, we set the probability of staying in a dynamic to 0.98 for each 2 ms time step. Because the probability of changing dynamics follows a geometric distribution, this corresponds to an expected duration of 100 ms for staying in a particular dynamic. Importantly, the data drive the dynamic, so even if the probability of a particular dynamic is high, data providing clear evidence for a change in dynamic can, within the time frame of ~10 ms, drive a change to a different dynamic.

@@

As mentioned above, the model is robust to the choice of probability of staying in a state. To demonstrate this we decoded the spiking sequence in Figure \ref{1}A with different probabilities of staying in the same state versus switching to another state (Figure 1-Supplemental Figure 1). We found that for a large range of plausible probabilities of staying in one of the dynamics (0.9 - 0.9999), the model still correctly identified the dynamic with high probability.

@@

Our success with simulated data inspired a further validation of the model on real hippocampal data. We fit the place fields of 31 cells recorded in hippocampus while the rat was performing a spatial alternation task on a W-shaped track, and then applied the decoding algorithm to a SWR with sequential population activity (Figure \ref{2}A, top panel). As expected, we observed that the probability of being in the continuous dynamic is high throughout this SWR, but the probability of being in a stationary state was noticeable at the beginning and end of the SWR (Figure \ref{2}A, middle panel). Using our speed classification scheme, this means that the speed of the replay starts slower---as a mixture of continuous and stationary dynamics---and then speeds up and slows down again. This is also evident in the posterior probability of linear position over time. This shows that the replay travels down the center arm and up the right arm (Figure \ref{2}A, bottom panel). We can also see this when we project the maximum of the posterior of this trajectory to 2D (Figure \ref{2}C) to better see the spatial trajectory. Moreover, when we apply the same model using the 2D position of the animal, we get a similar result as when we used the 1D linearized position (Figure 2-Supplemental 1A).

One of our criteria for a more optimal method is that it use all of the available spiking data. Using only clustered spikes discards all of the spike events that cannot be uniquely assigned to a putative single neuron, substantially reducing the quality of the resulting decoding. Additionally, spike sorting is not necessary to recover the underlying neural population dynamics \cite{TrautmannAccurateEstimationNeural2019}. We therefore adopted a "clusterless" approach which directly relates spikes and their multiunit spike waveform features to position without spike sorting, which we call clusterless decoding (see Methods). Clusterless decoding previously has been used to successfully identify theta sequences and replay sequences in the hippocampus (\cite{KloostermanBayesiandecodingusing2014, ChenTransductiveneuraldecoding2012,DengRapidclassificationhippocampal2016, KayConstantSubsecondCycling2020}). Applying a clusterless decoder to the same SWR, we get similar classification of the sequence (Figure \ref{2}B, D), both with 1D linearized position and 2D position (Figure 2-Supplemental 1B). The spatial extent of the event is longer and the estimate of the posterior probability of position is narrower for the clusterless model, however. This reflects the clusterless model's access to a larger pool of data that provides more information about the extent of the event, and more certainty in the latent position and the dynamic (Figure \ref{2}D vs C).

@@

After testing our model on a single SWR, we applied our decoding algorithm to hippocampal recordings from 10 animals performing the W-track spatial alternation task (tetrodes = [10, 24], brain areas = [CA1, CA2, CA3]; some data previously used in \cite{KarlssonAwakereplayremote2009, KayConstantSubsecondCycling2020}). One major goal of our work was to assess the overall prevalence of spatial content across SWRs, so we detected SWRs using a relatively permissive threshold (see Methods) to ensure that we were included both the rare large events as well as the much more common small events.

As expected, and providing further support for the validity of the method, we observed many replays that were classified as continuous throughout the entirety of SWR (Figure 2-Supplemental 2A-C). However, we also observed many events with spatially coherent content that did not have this structure. For example, there were trajectories that started in one direction and reversed back to the original position (Figure \ref{3}A, Figure 3-Supplemental 1B), trajectories that remained fixed in one position (Figure \ref{3}B), trajectories that jumped between arms and between dynamics (Figure \ref{3}C-D, Figure3-Supplemental 1A, E, F), and trajectories that were spatially incoherent throughout the SWR (Figure3-Supplemental1D).

Using our 0.8 threshold, we were able to classify 89\% ( 23382 of 26160) of SWRs as containing at least one of the five dynamics. To ensure that this depended on the spatial receptive fields of cells in the hippocampus, we resampled position with replacement for two epochs, shuffling the relationship between spiking and spatial information. We then decoded the same SWR events containing the same spikes. Only 9\% of the SWRs were classified in the shuffled datasets, a value that was significantly less than that seen for the real data (p=0.02 for epoch 1, p=0.02 for epoch 2, Figure 3-Supplemental 2).

Previous work focusing on spatially sequential replay reported that only a minority of events contain sequential spatial content (Karlsson and Frank, Foster and Wilson Davidson et. al. etc. . We therefore asked what fraction of classified events contained spatially coherent content in any form, defined as a SWR containing any times with stationary, stationary-continuous mixture, or continuous dynamics (see Methods). We found that 22478 of 23382 or 96\% of classified SWRs (85\% of all SWRs) included spatially coherent structure . We then asked what fraction of events contained spatially incoherent content, defined as a SWR containing any times with fragmented-continuous mixture or fragmented dynamics. We found that only 3353 of 23382 or 14\% of classified SWRs had time points with spatially incoherent structure. This was consistent across animals (Figure \ref{3}F).

To better compare our findings to previous work, we quantified the percentage of classified SWRs that contained continuous content, as would typically be analyzed when using the Bayesian decoder. Here our findings were consistent with previous reports: in our datasets 4511 of 23382 or 19\% of classified SWRs had some continuous content (Figure \ref{3}G). Thus, focusing on only continuous content misses a large fraction of events where there is evidence for spatially coherent, but not continuous content.

@@

Our results indicate that most events with coherent spatial content are not dominated by the continuous dynamic, and thus correspond to trajectories that move relatively slowly compared to their purely continuous counterparts. We therefore examined these events in more detail. We first noted that most of the SWRs (16454 of 23382 classified SWRs or 70\%) were well described by a single dynamic. Surprisingly, the most common SWRs were categorized exclusively as stationary-continuous mixtures (9860 of 23382 classified SWRs or 42\%, Figure \ref{5}A). These events contain representations that move at rates more consistent with real-world movement (Fig. 1G), but were slightly shorter, on average, than events with continuous dynamics (median duration: stationary-continuous-mixture 72 ms vs. continuous 95 ms, Figure \ref{5}B). Nonetheless, both these slow-moving events and continuous events represented locations that were some distance away from the animal (mean trajectory distance from animal's position: stationary-continuous-mixture 51 cm vs. continuous 42 cm, Figure \ref{5}C). This indicates that the trajectories of these SWRs, like ones that have faster continuous dynamics, do not represent the animal's actual position and cannot be explained by erroneously detected SWRs.

@@

Previous work reported that at most half of SWRs or high-population-activity events contained replay (REFS from above). This could lead one to the assumption that most events are not spatially coherent or meaningful. Our results indicate that this is not the case: in our dataset the large majority of SWRs (85\%) contained spatially coherent content, defined as content with some combination of stationary or continuous dynamics. These events included decoded representations that were highly spatial concentrated, confirming their spatial coherence. This was in stark contrast to unclassified events or events that included fragmented dynamics, which were associated with highly distributed decoded spatial representations.

Importantly, the spatially coherent events most often expressed a dynamic that was consistent with real-world movement speeds of <1 m/s. The next most common category were stationary events (Cite Yu etl al, Farooq et. al, perhaps Dupret et. al 2009?) that activated a representation of a single location, while the classical "replay" events made up only about ~19\% of classified events.

These results challenge the long-held notion that hippocampal replay trajectories necessarily proceed at many times the speed of the animal \cite{NadasdyReplayTimeCompression1999, LeeMemorySequentialExperience2002, DavidsonHippocampalReplayExtended2009}. Instead there is a much richer variety of speeds the animal can use to retrieve information about prior experience or generate possible future events, including slower moving representations that activate a spatially compact set of nearby locations and stationary representations of single locations. Interestingly, Farooq and Dragoi \cite{FarooqEmergencepreconfiguredplastic2019} reported that these stationary representations were most prevalent in juvenile rats and decreased to "chance" levels when the rats reached adulthood. Here, we show that they do not disappear, but are a very common occurrence with adult rats. Here we note that we chose to include a stationary distribution to explicitly evaluate the prevalence of stationary events in adult animals, but in principle a model with just continuous and fragmented dynamics could capture these events as well, as the continuous dynamic includes movement speeds of 0 m/s.

@@

Second, we used clusterless decoding, which enabled us to take advantage of information from neurons farther away from the tetrodes and not just those that could be cleanly and somewhat subjectively separated (REFS). Third, we built an explicit, but flexible model that allowed us to identify not just one dynamic, but multiple dynamics consistent with different speeds of motion. Fourth, our model avoids assumptions about temporal bins and does not make assumptions about the linear progression of events. Instead, the model can use any size of bin and allows for movement in any direction. Finally, we used an acausal estimator of latent position, which allowed us to identify the specific timing of transitions in dynamics at a fine timescale.

@@

\subsection*{Behavioral Task}
All animals performed a W-track spatial alternation task, which is described in detail in \cite{KarlssonAwakereplayremote2009}. In brief, each day, animals alternated between 20 minute rest epochs in a rest box and 15 minutes run epochs in the W-track. On the W-track, animals were rewarded at the ends of an arm when that arm was the next correct arm in the sequence. Two rules determined the next correct arm. If the animal was in an outer arm, it had to run to the center arm. If the animal was in the center arm, it had to run from the center arm to the least recently visited outer arm. Animals were free to choose any arm at any time Only run epochs with at least 9 putative hippocampal pyramidal cells that fired at least 100 spikes were included in the analysis.

@@

\begin{figure}%[tbhp]
\centering
\includegraphics[width=0.80\linewidth]{figures/Figure3-supplemental2/Figure3_v1-supplemental2}
\caption{Comparison of real vs. position shuffled data for two epochs from different animals. Line represents the percent of SWR classified in that epoch for real data. The histogram represents the distribution after 50 shuffles of the position data. Position data was shuffled by resampling with replacement from the set of all observed positions in that epoch, destroying position information but preserving spiking timing.}
\label{Figure3-Figure supplement 2}
\end{figure}

>>>>

Mental events may not occur in the same time span as actual events: memories can be retrieved and used in less time than was required for the original experience, and mental simulations can span a series of actions more quickly than would be required to perform those actions. Consistent with this, generative representations can unfold rapidly \cite{NadasdyReplayTimeCompression1999, LeeMemorySequentialExperience2002, DavidsonHippocampalReplayExtended2009, KarlssonAwakereplayremote2009}. As a result, decoding generative representations often involves making assumptions about their timescale and temporal structure, and these assumptions can limit the interpretation of the data.

@@

Identifying these events typically involves multiple steps and assumptions about the nature of the event. First, an encoding model is constructed based on spiking activity during movement, most often using only spikes that have been clustered as single units (putative single neurons). Then, a subset of SWRs or events with high activity levels are selected based on an threshold for event size chosen by the experimenter \cite{FosterReversereplaybehavioural2006, DibaForwardreversehippocampal2007, KarlssonAwakereplayremote2009, StellaHippocampalReactivationRandom2019}. The spikes within these events decode the probability of position in overlapping or non-overlapping time bins whose size is also chosen by the experimenter. Finally, the most commonly used approach to detecting sequential replay involves fitting a line to the resulting set of probability distributions, which instantiates the assumption that the representations progress at a constant speed \cite{FosterReversereplaybehavioural2006, DibaForwardreversehippocampal2007, KarlssonAwakereplayremote2009}. A statistical test is then used to determining whether the fit is better than the fit to shuffled versions of the data.

While this approach identifies constant speed events, it does not consider events that are rejected by the statistical test, and the use of a fixed size temporal bin acts as a boxcar smoother that limits the potential movement speeds of the representation. The linear fit is also problematic, because it has the potential to reject real events that do not move at constant speeds. Approaches that focus on the order of cell activity within each even \cite{LeeMemorySequentialExperience2002, GuptaHippocampalReplayNot2010} allow for a relaxation of that linear assumption, but replace by either ignoring all but the first spike from each cell or some arbitrarily smoothed version of the spike train. These approaches also do not provide information about the dynamics of the underlying spatial representation. Moreover, there is now evidence that a subset of these events have stationary representations of a single location, which both approaches exclude. In addition, these methods identify only a subset of events as containing content (typically somewhere between 10 and 45\% of the chosen events), and thereby provide little insight into the remaining majority of the events.

@@

\section*{Results}
\begin{figure*}%[tbhp]
\centering
\includegraphics[width=0.80\linewidth]{figures/Figure1/Figure1_v4}
\caption{Components of the state space model and demonstration on simulated data. \textbf{(A)} 19 simulated place cells that exhibit three different sequence patterns. For the first third of time, one cell fires repeatedly, representing one stationary location. For the second third of time, the cells fire in sequence, representing a smooth continuous trajectory through space. For the last third cells fire in an incoherent pattern, representing a non-smooth fragmented trajectory through space. \textbf{(B)} Like the Bayesian Decoder, the state space model uses estimates of cells' place fields from when the animal is moving and combines them with the observed spikes in (A) to compute the likelihood of position for each time step. \textbf{(C)} This is combined with an explicit model of each dynamic state--which determines how latent position can change based on the posterior from the previous time step. In the stationary state, this is modeled as the identity matrix because the position remains the same. In the continuous state, the dynamics are modeled as a Gaussian random walk, meaning the next possible position is likely to be spatially close. In the fragmented state, the dynamics are modeled as uniform, meaning the next position is equally likely to be anywhere. Note that the dynamics displayed only reflect the within-state dynamics. \textbf{(D)} The probability of remaining in a particular state versus switching to another state. This is modeled as having a high probability of remaining in a particular state with a small probability of switching to one of the other dynamics at each time step. \textbf{(F)} The state space model uses the components in A-D over all time to decode the joint probability of latent position and dynamic state. The joint probability can be summarized by marginalizing over latent position (left panel) to get the probability of each dynamic state over time (lines). This can also be used to get an estimate of latent position over time (right panel). \textbf{(G)} The probability of each state given a particular constant speed trajectory. Each speed corresponds to a simulated spiking sequence at that speed. The lines correspond to the average probability of that state over the sequence. Dotted line represents the 0.8 classification threshold we use to classify each range of speeds and shaded regions correspond to the classification we have given the range of speeds.
}
\label{1}
\end{figure*}
\subsection*{Overview of the model}
We begin with an application of the model to simulated data, both to validate the model and to provide intuition (Figure \ref{1}). We simulated 19 Poisson spiking cells with Gaussian place fields on a 180 cm virtual linear track. Each place field has a 36 cm variance and a 15 Hz peak firing rate, which is spaced every 10 cm along the virtual track. We then apply our decoding algorithm to the spiking sequence in Figure \ref{1}A. For the first third of this sequence, a single place cell fires repeatedly, representing a single location. For the middle third of the sequence, the cells fire in sequential spatial order, representing a fast moving trajectory. For the last third, the cells fire in a incoherent spatial order. These firing patterns represent three different types of sequence dynamics that could be expressed during replay events, which we call stationary, continuous, and fragmented, respectively. The goal of our model is to characterize SWRs in terms of a mixture of these three dynamics at every time point.

Decoding the spiking sequence dynamics requires specifying two elements: the data model---how the spikes relate to position---and the movement dynamics model---how the position can change over time. For the data model, our decoder is the same as the Bayesian decoder. We compute an estimate of how each cell's firing rate varies over position (the place field, Figure \ref{1}B). This is used during decoding to compute the Poisson likelihood of position over time given the spiking sequence of interest. In contrast to the Bayesian decoder, we can use arbitrarily small time bins (in our case 2 ms vs. 20 ms or more) because we are able to take advantage of the prior placed on the dynamics by the state space model. This allows us to detect changes on smaller time scales than would be possible with the Bayesian decoder.

Next, we specify a movement dynamics model for how latent position---the "mental" position of the animal represented by the cells---evolves over time (Figure \ref{1}C). We do this by defining a movement state transition matrix, which defines how the latent position can change from the previous time step. Previous findings suggest three distinct dynamics: stationary \cite{JaiDistincthippocampalcorticalmemory2017, FarooqEmergencepreconfiguredplastic2019}, smoothly moving \cite{DavidsonHippocampalReplayExtended2009}, and fragmented / jumpy \cite{PfeifferAutoassociativedynamicsgeneration2015}, which could correspond to both spatially incoherent representations and representations that jump from one position to another. We therefore developed state transition models to capture each of these dynamics.

@@

Finally, we specify how likely each dynamic is to persist in time versus changing to another dynamic (Figure \ref{1}D). In order to be conservative, we assume that each dynamic is likely to dominate for ~100 ms, which is approximately the duration of the SWR. There is, however, a small probability of switching to one of the other dynamics (note that below we show that our results are relatively insensitive to this parameter). Accordingly, we set the probability of staying in a dynamic to 0.98 for each 2 ms time step. Because the probability of changing dynamics follows a geometric distribution, this corresponds to an expected duration of 100 ms for staying in a particular dynamic. Importantly, the data drive the dynamic, so even if the probability of a particular dynamic is high, data providing clear evidence for a change in dynamic can, within the time frame of ~10 ms, drive a change to a different dynamic.

@@

As mentioned above, the model is robust to the choice of probability of staying in a state. To demonstrate this we decoded the spiking sequence in Figure \ref{1}A with different probabilities of staying in the same state versus switching to another state (Figure \ref{fig:Figure1-Figure supplement 1}). We found that for a large range of plausible probabilities of staying in one of the dynamics (0.9 - 0.9999), the model still correctly identified the dynamic with high probability.

@@

Our success with simulated data inspired a further validation of the model on real hippocampal data. We fit the place fields of 31 cells recorded in hippocampus while the rat was performing a spatial alternation task on a W-shaped track, and then applied the decoding algorithm to a SWR with sequential population activity (Figure \ref{2}A, top panel). As expected, we observed that the probability of being in the continuous dynamic is high throughout this SWR, but the probability of being in a stationary state was noticeable at the beginning and end of the SWR (Figure \ref{2}A, middle panel). Using our speed classification scheme, this means that the speed of the replay starts slower---as a mixture of continuous and stationary dynamics---and then speeds up and slows down again. This is also evident in the posterior probability of linear position over time. This shows that the replay travels down the center arm and up the right arm (Figure \ref{2}A, bottom panel). We can also see this when we project the maximum of the posterior of this trajectory to 2D (Figure \ref{2}C) to better see the spatial trajectory. Moreover, when we apply the same model using the 2D position of the animal, we get a similar result as when we used the 1D linearized position (Figure \ref{fig:Figure2-Figure supplement 1}A).

One of our criteria for a more optimal method is that it use all of the available spiking data. Using only clustered spikes discards all of the spike events that cannot be uniquely assigned to a putative single neuron, substantially reducing the quality of the resulting decoding. Additionally, spike sorting is not necessary to recover the underlying neural population dynamics \cite{TrautmannAccurateEstimationNeural2019}. We therefore adopted a "clusterless" approach which directly relates spikes and their multiunit spike waveform features to position without spike sorting, which we call clusterless decoding (see Methods). Clusterless decoding previously has been used to successfully identify theta sequences and replay sequences in the hippocampus (\cite{KloostermanBayesiandecodingusing2014, ChenTransductiveneuraldecoding2012,DengRapidclassificationhippocampal2016, KayConstantSubsecondCycling2020}). Applying a clusterless decoder to the same SWR, we get similar classification of the sequence (Figure \ref{2}B, D), both with 1D linearized position and 2D position (Figure \ref{fig:Figure2-Figure supplement 1}B). The spatial extent of the event is longer and the estimate of the posterior probability of position is narrower for the clusterless model, however. This reflects the clusterless model's access to a larger pool of data that provides more information about the extent of the event, and more certainty in the latent position and the dynamic (Figure \ref{2}D vs C).

@@

After testing our model on a single SWR, we applied our decoding algorithm to hippocampal recordings from 10 animals performing the W-track spatial alternation task (\# tetrodes range: [10, 24], brain areas = [CA1, CA2, CA3]; some data previously used in \cite{KarlssonAwakereplayremote2009, KayConstantSubsecondCycling2020, CarrTransientSlowGamma2012}). One major goal of our work was to assess the overall prevalence of spatial content across SWRs, so we detected SWRs using a relatively permissive threshold (see Methods) to ensure that we were included both the rare large events as well as the much more common small events.

As expected, and providing further support for the validity of the method, we observed many replays that were classified as continuous throughout the entirety of SWR (Figure \ref{fig:Figure2-Figure supplement 2}A-C). However, we also observed many events with spatially coherent content that did not have this structure. For example, there were trajectories that started in one direction and reversed back to the original position (Figure \ref{3}A, Figure \ref{fig:Figure3-Figure supplement 1}1), trajectories that remained fixed in one position (Figure \ref{3}B), trajectories that jumped between arms and between dynamics (Figure \ref{3}C-D, Figure \ref{fig:Figure3-Figure supplement 1}A, E, F), and trajectories that were spatially incoherent throughout the SWR (Figure \ref{fig:Figure3-Figure supplement 1}D).

Using our 0.8 threshold, we were able to classify 89\% ( 23382 of 26160) of SWRs as containing at least one of the five dynamics. To ensure that this depended on the spatial receptive fields of cells in the hippocampus, we resampled position with replacement for two epochs, shuffling the relationship between spiking and spatial information. We then decoded the same SWR events containing the same spikes. Only 9\% of the SWRs were classified in the shuffled datasets, a value that was significantly less than that seen for the real data (p=0.02 for epoch 1, p=0.02 for epoch 2, Figure \ref{fig:Figure3-Figure supplement 2}).

Previous work focusing on spatially sequential replay reported that only a minority of events contain sequential spatial content \cite{KarlssonAwakereplayremote2009, FosterReversereplaybehavioural2006, DavidsonHippocampalReplayExtended2009}. We therefore asked what fraction of classified events contained spatially coherent content in any form, defined as a SWR containing any times with stationary, stationary-continuous mixture, or continuous dynamics (see Methods). We found that 22478 of 23382 or 96\% of classified SWRs (85\% of all SWRs) included spatially coherent structure. We then asked what fraction of events contained spatially incoherent content, defined as a SWR containing any times with fragmented-continuous mixture or fragmented dynamics. We found that only 3353 of 23382 or 14\% of classified SWRs had time points with spatially incoherent structure. This was consistent across animals (Figure \ref{3}F).

To better compare our findings to previous work, we quantified the percentage of classified SWRs that contained continuous content, as would typically be analyzed when using the Bayesian decoder. Here our findings were consistent with previous reports: in our datasets 4511 of 23382 or 19\% of classified SWRs had some continuous content (Figure \ref{3}G, 17\% of all SWRs). Thus, focusing on only continuous content misses a large fraction of events where there is evidence for spatially coherent, but not continuous content.

@@

Our results indicate that most events with coherent spatial content are not dominated by the continuous dynamic, and thus correspond to trajectories that move relatively slowly compared to their purely continuous counterparts. We therefore examined these events in more detail. We first noted that most of the SWRs (16454 of 23382 classified SWRs or 70\%) were well described by a single dynamic. Surprisingly, the most common SWRs were categorized exclusively as stationary-continuous mixtures (9860 of 23382 classified SWRs or 42\%, Figure \ref{5}A). These events contain representations that move at rates more consistent with real-world movement (Figure \ref{1}G), but were slightly shorter, on average, than events with continuous dynamics (median duration: stationary-continuous-mixture 72 ms vs. continuous 95 ms, Figure \ref{5}B). Nonetheless, both these slow-moving events and continuous events represented locations that were some distance away from the animal (mean trajectory distance from animal's position: stationary-continuous-mixture 51 cm vs. continuous 42 cm, Figure \ref{5}C). This indicates that the trajectories of these SWRs, like ones that have faster continuous dynamics, do not represent the animal's actual position and cannot be explained by erroneously detected SWRs.

@@

Previous work reported that at most half of SWRs or high-population-activity events contained replay \cite{KarlssonAwakereplayremote2009, FosterReversereplaybehavioural2006, DavidsonHippocampalReplayExtended2009}. This could lead one to the assumption that most events are not spatially coherent or meaningful. Our results indicate that this is not the case: in our dataset the large majority of SWRs (85\%) contained spatially coherent content, defined as content with some combination of stationary or continuous dynamics. These events included decoded representations that were highly spatial concentrated, confirming their spatial coherence. This was in stark contrast to unclassified events or events that included fragmented dynamics, which were associated with highly distributed decoded spatial representations.

Importantly, the spatially coherent events most often expressed a dynamic that was consistent with real-world movement speeds of <1 m/s. The next most common category were stationary events \cite{JaiDistincthippocampalcorticalmemory2017, FarooqEmergencepreconfiguredplastic2019} that activated a representation of a single location, while the classical "replay" events made up only about ~19\% of classified events.

These results challenge the long-held notion that hippocampal replay trajectories necessarily proceed at many times the speed of the animal \cite{NadasdyReplayTimeCompression1999, LeeMemorySequentialExperience2002, DavidsonHippocampalReplayExtended2009}. Instead there is a much richer variety of speeds the animal can use to retrieve information about prior experience or generate possible future events, including slower moving representations that activate a spatially compact set of nearby locations and stationary representations of single locations. Interestingly, Farooq and Dragoi \cite{FarooqEmergencepreconfiguredplastic2019} reported that these stationary representations were most prevalent in juvenile rats and decreased to "chance" levels when the rats reached adulthood. Here, we show that they do not disappear, but are a very common occurrence with adult rats. We note that we chose to include a stationary distribution to explicitly evaluate the prevalence of stationary events in adult animals, but in principle a model with just continuous and fragmented dynamics could capture these events as well, as the continuous dynamic includes movement speeds of 0 m/s.

@@

Second, we used clusterless decoding, which enabled us to take advantage of information from neurons farther away from the tetrodes and not just those that could be cleanly and somewhat subjectively separated \cite{ChenBayesiannonparametricmethods2016, KloostermanBayesiandecodingusing2014, DengRapidclassificationhippocampal2016}. Third, we built an explicit, but flexible model that allowed us to identify not just one dynamic, but multiple dynamics consistent with different speeds of motion. Fourth, our model avoids assumptions about temporal bins and does not make assumptions about the linear progression of events. Instead, the model can use any size of bin and allows for movement in any direction. Finally, we used an acausal estimator of latent position, which allowed us to identify the specific timing of transitions in dynamics at a fine timescale.

@@

\subsection*{Behavioral Task}
All animals performed a W-track spatial alternation task, which is described in detail in \cite{KarlssonAwakereplayremote2009}. In brief, each day, animals alternated between 20 minute rest epochs in a rest box and 15 minutes run epochs in the W-track. On the W-track, animals were rewarded at the ends of an arm when that arm was the next correct arm in the sequence. Two rules determined the next correct arm. If the animal was in an outer arm, it had to run to the center arm. If the animal was in the center arm, it had to run from the center arm to the least recently visited outer arm. Animals were free to choose any arm at any time. Only run epochs with at least 9 putative hippocampal pyramidal cells that fired at least 100 spikes were included in the analysis.

@@

\begin{figure}%[tbhp]
\centering
\includegraphics[width=0.80\linewidth]{figures/Figure3-supplemental2/Figure3_v1-supplemental2}
\caption{Comparison of real vs. position shuffled data for two epochs from different animals. Line represents the percent of SWR classified in that epoch for real data. The histogram represents the distribution after 50 shuffles of the position data. Position data was shuffled by resampling with replacement from the set of all observed positions in that epoch, destroying position information but preserving spiking timing.}
\label{fig:Figure3-Figure supplement 2}
\end{figure}
